---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message=FALSE,
  warning=FALSE,
  results="hide"
)
```

Welcome! Let's get you started with bbsBayes2.

First we'll make sure we have the right software installed, we'll fetch
the BBS survey data, and then we'll run through some example
workflows.

## Install [cmdstanr](https://mc-stan.org/cmdstanr/)

Because bbsBayes2 uses Stan to run the Bayesian models, we need to make
sure we have cmdstanr and cmdstan both installed.

It is recommended to run this is a fresh R session or restart your
current session

```{r, eval = FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", 
                                       getOption("repos")))
```

Now we should be able to use cmdstanr to install cmdstan

```{r, eval = FALSE}
cmdstanr::install_cmdstan()
```

Let's check that everything went as planned

```{r, eval = TRUE}
cmdstanr::check_cmdstan_toolchain()
```

> Problems? Check out cmdstanr's vignette on [Getting
> Started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)

### Special note for Windows users

We strongly recommend that you install Linux on your Windows machine and
take advantage of `cmdstanr` functions that run `Stan` models in Linux.
This will likely cut the MCMC run-times by 30-50% for the `bbsBayes2`
models. Installing [Windows Subsystem for Linux
(WSL)](https://learn.microsoft.com/en-us/windows/wsl/install) is a small
hassle, but only needs to be done once. Follow the directions at the
above link.

Once the WSL installation is complete, re-install `cmdstan` using `cmdstanr::install_cmdstan(overwrite = TRUE, wsl = TRUE)`.
Now, everytime you run a model using `cmdstan` (and therefore anytime you run a model using `bbsBayes2`), it will use the Linux installation to run Stan. It's seamless and you'll be very thankful you did it once it's done.

## Download BBS data

Now we'll fetch the BBS data using functions included in bbsBayes2.

> Note that most bbsBayes2 functionality can be explored **without**
> downloading BBS data by using the included sample data. Specify
> `sample_data = TRUE` in the first `stratify()` step (see the next
> section).

First we'll load the package.

```{r, message = FALSE}
library(bbsBayes2)
```

There are (as of June 2023) four types of BBS data that can be downloaded:

-   Two levels `state` and `stop` (only `state` works with bbsBayes2
    models)
-   Two releases `2020` and `2022` (more options will be added as annual releases occur)

Unless you have a specific reason to need a different version, the
defaults (level `state` and the most recent release) is what you want.
However you can download all data sets and specify which one you wish to
use in the `stratify()` step.

### A note about BBS release names: 
The releases are named for the year in which the annual dataset was released after going through QA-QC. So the `2022` release contains all surveys conducted up to and including the 2021 season.
There is no release for 2021, because the BBS was cancelled during the COVID lockdowns of spring 2020 and there are no 2020 BBS data.

```{r, eval = FALSE}
fetch_bbs_data()                  # Default
fetch_bbs_data(release = "2020")  # Specify a different release
```

```{r, echo = FALSE, eval = !have_bbs_data()}
fetch_bbs_data()
```

## Workflow

Now that you have cmdstanr installed and the BBS data downloaded, we'll
walk through a general workflow for modelling species trends with
bbsBayes2.

### Stratify the data

The first step in any bbsBayes2 analysis is to stratify the data. In
this step we choose a stratification type as well as a species to
explore.

```{r, eval = have_bbs_data()}
s <- stratify(by = "bbs_usgs", species = "Scissor-tailed Flycatcher")
```

We can also play around with the included sample data (Pacific Wrens)

```{r}
s <- stratify(by = "bbs_cws", sample_data = TRUE) # Only Pacific Wren
```

Stratifications included in bbsBayes2 are
`r paste0(names(bbs_strata), collapse = ", ")`. See the articles on
[stratifications](articles/stratification) and [custom
stratifications](articles/custom_stratification) for more details and
examples.

Available species can be explore using the `search_species()` function.
Note: if you're just messing around with the package we recommend using
species with relatively small ranges (i.e., relatively few data) such as the Hepatic Tanager, Pacific Wren, Scissor-tailed Flycatcher and Cerulean Warbler.

### Prepare the data

Once we have stratified the data, we can now prepare it for use in a
model. In this step data will be filtered to omit routes with too few
samples, etc. See `prepare_data()` for more details on how you can
customize this step.

```{r}
p <- prepare_data(s)
```

### Prepare the model

Next we will prepare the model parameters and initialization values. See
`prepare_model()` for more details on how you can customize this step.

```{r}
md <- prepare_model(p, model = "first_diff")
```

### Run model

Now we can run the model.

The default `iter_sampling` and `iter_warmup` are 1000 and the default
`chains` is 4. In the interest of speed for this example, we are using
much lower values, but note that this almost certainly will result in
problems with our model.

```{r, eval = FALSE}
m <- run_model(md, iter_sampling = 100, iter_warmup = 500, chains = 2)
```


### A note on spatial models

For spatial models there's two additional steps. You stratify and
prepare the data as in the previous example, but you also prepare the
map and the spatial data. An example is below.

```{r, eval=FALSE}
s <- stratify(by = "bbs_usgs", species="Scissor-tailed Flycatcher") 
p <- prepare_data(s)

#now the additional steps
#load a map
map<-load_map(stratify_by = "bbs_usgs")
#prepare the spatial data
sp<-prepare_spatial(p,map)

#then prepare the model with the spatial output
mod <- prepare_model(sp, model = "gamye",model_variant = "spatial")

#then run the model as before
m <- run_model(mod)

#optionally, save the model output as an .rds file
saveRDS(m,"output/4430_gamye_spatial.rds")
```

### Explore the model outputs

If you would prefer to skip the model fitting steps for now, you can download a fitted model object (the output of `run_model()` function) and test out the remaining package features.

The outputs of the collection of functions required to fit a model are cumulative: each one retains the metadata from the previous step. As a result, the saved object from the `run_model()` function is a large list that includes the `cmdstanr` posterior samples object from the model fitting process, as well as all of the data and metadata necessary to understand and replicate the choices made to fit the model.  

```{r}
m<-readRDS("output/4430_gamye_spatial.rds")
```

First we will generate indices according to different regional
categories. By default these are `continent` and `stratum`. Indices (annual indices of relative abundance) represent mean predicted annual counts of the species in a given region, on an average route, by an average observer. The pattern in the time-series of these annual indices for a given region represent the estimated population trajectory.  

```{r}
i <- generate_indices(model_output = m)
```

We can explore or extract these indices for saving as an external file, by accessing the `indices` item in
the list. 

```{r}
i[["indices"]]
```

We can also plot these indices.

```{r}
p <- plot_indices(indices = i, add_observed_means = TRUE)
```

Note that we get one plot for each region and regional category, in this
case that means one plot for the continent, and one for each stratum.

```{r}
names(p)
```

We can plot them individually by pulling a plot out of the list

```{r}
p[["continent"]]
```

Next we can calculate the specific trends based on these indices

```{r}
t <- generate_trends(i)
```

We can plot these on a map

```{r}
plot_map(t)
```


###Example models 

A model with a suitable number of iterations takes a
long time to run (the barn swallow model below took 54 hours)! Below are
three full models run with 1000 iterations and four chains. Use these to
test out some of the features!

Download the Barn Swallow GAMYE spatial model:
<https://github.com/L-Daly/bbsbae/releases/tag/trial> click
BASW_spatial_gamye.zip. Download the Hepatic Tanager GAMYE model, by
clicking "Download" here: \<
<https://github.com/L-Daly/bbsbae/blob/25eeb0f974bcdf832b8375422a9ac57c046c62c2/data/HETA_bbs_usgs_gamye.rds>
. Download the Hepatic Tanager model stratified by latitude/longitude by
clicking "Download" here:
<https://github.com/L-Daly/bbsbae/blob/25eeb0f974bcdf832b8375422a9ac57c046c62c2/data/HETA_latlong_gam.rds>.

An example of the output from applying the spatial gamye model to Barn Swallow data.

```{r}
#First is a spatial GAMYE model for the Barn Swallow
# library(readr)
# BARS <- read_rds(unzip(
#   "BASW_spatial_gamye.zip",
#   "Barn_Swallow_gamye_spatial.rds"))

BARS <- readRDS("output/Barn_Swallow_gamye_spatial.rds")
#investigate model meta data
BARS$meta_data
#length of run-time in seconds
BARS$model_fit$time()
#length of run-time in hours
BARS$model_fit$time()$total/3600

BARS_indices <- generate_indices(BARS)

BARS_index_plots <- plot_indices(BARS_indices,
                                 add_observed_means = TRUE,
                                 add_number_routes = TRUE)

BARS_index_plots[["continent"]]
```

Above, we've plotted the full annual indices (i.e., the population trajectory) from the gamye model, which include the smooth component and the annual fluctuations. 
We can also generate smooth annual indices representing the spline-based smooth component of the population trajectory, after removing the random annual fluctuations. These smooth trajectories are useful for trend estimates that vary less from year-to-year.

```{r}
BARS_smooth_indices <- generate_indices(BARS,
                                 alternate_n = "n_smooth")
BARS_trends <- generate_trends(BARS_smooth_indices)
BARS_trend_map <- plot_map(BARS_trends)
BARS_trend_map

```

We also have results for Hepatic Tanager from two different models and using different spatial stratifications. 

```{r}

#Next we have a Hepatic Tanager gamye model
HETA_gamye<- readRDS("output/HETA_bbs_usgs_gamye.rds")
HETA_gamye_indices <- generate_indices(HETA_gamye)
HETA_gamye_trends <- generate_trends(HETA_gamye_indices)
HETA_gamye_map <- plot_map(HETA_gamye_trends)
HETA_gamye_map

```


```{r}

#lastly we have a Hepatic Tanager GAM model, stratified to latitude and longitude

HETA_gam_latlong <- readRDS("output/HETA_latlong_gam.rds")
HETA_gam_latlong_indices <- generate_indices(HETA_gam_latlong)
HETA_gam_latlong_trends <- generate_trends(HETA_gam_latlong_indices)
HETA_gam_latlong_map <- plot_map(HETA_gam_latlong_trends)
HETA_gam_latlong_map
```



```{r, include = FALSE}
unlink(list.files(pattern = "^BBS_STAN_(\\w|\\d|-)+.(csv|rds)"))
```

And there you have it!

Check out the other [articles](/index.html) to explore more advanced
usage or the [function reference](../reference/) to see what functions
are available and how to use them in greater detail.
