---
title: "Models in bbsBayes2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models in bbsBayes2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}

library(bbsBayes2)
```

There are 9 types of models that can be run with bbsBayes2. For a quick overview
you can access the `bbs_models` data frame.

```{r}

bbs_models
```


For specifics on *how* to run the various models, see the articles on 
[model runs](./index.html#model-runs).

See the following for details regarding the *why* of these various models. 
As well as the final section on Advanced Modelling techniques. All models can be hierarchical, non-hierarchical, or spatial. Hierarchical models share information on species abundance and trend over time, so that each time periods' estimates are pulled closer to the mean for that species across time. Non-hierarchical models have no sharing of information, and consider time periods and spatial strata as independent. Spatial models are both hierarchical and spatial, sharing information over time and space. This means that each stratum's estimates will be pulled closer to the mean of each its neighbors and to the species mean across time.

## 1. First Difference Models

A first-difference model considers the time-series as a random-walk from the first year, so that the first-differences of the sequence of year-effects are random effects with mean = 0 and an estimated variance. This model has been described in Link et al. 2017 https://doi.org/10.1650/CONDOR-17-1.1


## 2. General Additive Models
The GAM models the time series as a semiparametric smooth using a Generalized Additive Model (GAM) structure. See Smith & Edwards, 2021 https://doi.org/10.1093/ornithapp/duaa065. 

## 3. General Additive Models Year Effect
The GAMYE includes the semiparametric smooth used in the gam option, but also includes random year-effect terms that track annual fluctuations around the smooth. This is the model that the Canadian Wildlife Service is now using for the annual status and trend estimates. See pre-print at Smith et al., 2023 https://doi.org/10.32942/X2088D.

## 4. Slope
The slope option estimates the time series as a log-linear regression with random year-effect terms that allow the trajectory to depart from the smooth regression line. It is the model used by the USGS and CWS to estimate BBS trends since 2011. The basic model was first described in 2002 (Link and Sauer 2002; https://doi.org/10.1890/0012-9658(2002)083[2832:AHAOPC]2.0.CO;2) and its application to the annual status and trend estimates is documented in Sauer and Link (2011; https://doi.org/10.1525/auk.2010.09220) and Smith et al. (2014; https://doi.org/10.22621/cfn.v128i2.1565).


## Advanced Modelling

### Cross-fold Validation

For cross fold validation, you'll need to define which observations are part 
of a test set and which are part of a training set.

To use the method implemented by bbsBayes2, we'll specify `calculate_cv` as
`TRUE` during the `prepare_model()` step. By default, this will prepare 10 test
sets (K-folds, `cv_k`) using `obs_n` as the grouping variable (`cv_fold_groups`)
and omitting groups with only single observations (`cv_omit_singles`).

MORE DETAILS ON HOW CHOSEN

```{r}
#a model of counts from 2009-2019
md <- stratify(by = "bbs_usgs", species = "Hepatic Tanager", quiet = TRUE) %>%
  prepare_data(min_year= 2009, max_year = 2019) %>%
  prepare_model("first_diff", calculate_cv = TRUE, cv_k = 3) 
```

Now our model data (`md`) has a list item called `folds` which contains the 
fold groups assigned to each observation. To use this, you'll need to loop your
model and specify the `k` iteration.

```{r}
#a quick model with too few iterations, not saving the example model
for(k in 1:3) {
  run_model(md, k = k, chains = 2, iter_warmup = 500, iter_sampling = 100,
            save_model=FALSE)
}
```

AND NOW...

### Customizing initial parameter values

To customize the initial parameter values you can create the model data with
`prepare_model()` and then either edit or overwrite the `init_values` item in
the list.

Here we create a relatively fast model with the minimum number of iterations, for 10 years, the minimum period of time that trends can be reasonably created.

```{r}
#Hepatic Tanager model for 2009-2019
md <- stratify(by = "bbs_usgs", species="Hepatic Tanager", quiet = TRUE) %>%
  prepare_data(min_year = 2009,max_year = 2019) %>%
  prepare_model(model = "first_diff", model_variant = "hier")
```

Make some changes to the `init_values` and pass the `model_data` to
`run_model()`.

```{r}
md[["init_values"]][["sdnoise"]] <- 0.5

m <- run_model(md, iter_warmup = 500, iter_sampling = 100, chains = 2,save_model=FALSE)
```

Note that you can supply different init values for each chain. If you supply one
set of init values (which is what `prepare_model()` outputs), `run_model()`, 
will duplicate them for each chain for you. If you supply a list of init values
such that the length is the same as the number of chains, your input will be 
used as is.

### Customize Stan Model File

Copy model file (and modify by hand in text editor)
```{r}
new_model <- copy_model_file("first_diff", "hier", tempdir())
new_model
```

Then give the file path of this new file to `prepare_model()`
```{r}
model_data <- stratify(by = "bbs_usgs", species="Hepatic Tanager", quiet = TRUE) %>%
  prepare_data(min_year = 2009,max_year = 2019) %>%
  prepare_model(model = "first_diff", model_variant = "hier",
                model_file = new_model)
```


```{r, include = FALSE}
unlink(list.files(pattern = "^BBS_STAN_(\\w|\\d|-)+.(csv|rds)"))
```
